# DIABLO : gene-miRNA network specific to breast cancer subtypes

#IMPORTS
```{r}
install.packages('stringr')
```

```{r}
install.packages("caTools")
```

```{r}
library(stringr)
```

```{r}
library('dplyr')
```

```{r}
library(caTools)
```

```{r}
library(mixOmics) 
```

#DATA LOAD

##Clinical Dataset
```{r}
#data.clinical_OR <- read.table("C:\\Users\\arati\\OneDrive - FCT NOVA\\PBCM\\Data_TGCA\\BRCA_clinicalData.txt",sep="\t",header=T)

data.clinical_OR <- read.table("C:\\Users\\USER~\\OneDrive - FCT NOVA\\PBCM\\Data_TGCA\\BRCA_clinicalData.txt",sep="\t",header=T)
```

```{r}
data.clinical_OR$sampleId<-str_replace_all(data.clinical_OR$sampleId, '-', '.')
```

##miRNA Dataset
```{r}
#data.mirna_OR<- read.table("C:\\Users\\arati\\OneDrive - FCT NOVA\\PBCM\\Data_TGCA\\BRCA_microRNA_expression.txt",sep="\t",header=T)

data.mirna_OR<- read.table("C:\\Users\\USER~\\OneDrive - FCT NOVA\\PBCM\\Data_TGCA\\BRCA_microRNA_expression.txt",sep="\t",header=T)
```

##mRNA Dataset
```{r}
#data.mrna_OR<- read.table("C:\\Users\\arati\\OneDrive - FCT NOVA\\PBCM\\Data_TGCA\\BRCA_mRNAs_expression.txt",sep="\t",header=T)

data.mrna_OR<- read.table("C:\\Users\\USER~\\OneDrive - FCT NOVA\\PBCM\\Data_TGCA\\BRCA_mRNAs_expression.txt",sep="\t",header=T)
```

#Data Preprocessing

##Remove "Normal-Like" samples

Clinical dataset clean
```{r}
data.clinical<- subset(data.clinical_OR, PAM50_SUBTYPE != 'Normal-like')
```

See "Normal-like" samples
```{r}
normal.like<- data.clinical_OR$sampleId[data.clinical_OR$PAM50_SUBTYPE=='Normal-like']
```

```{r}
normal.like
```

Clean miRNA dataset
```{r}
data.mirna<- data.mirna_OR[, !names(data.mirna_OR) %in% normal.like]
```

```{r}
intersect(names(data.mirna_OR), normal.like)
```
Clean mRNA dataset
```{r}
data.mrna<- data.mrna_OR[, !names(data.mrna_OR) %in% normal.like]
```

```{r}
intersect(names(data.mrna_OR), normal.like)
```

## Filter data with variance above 0.02

Variance filter - mRNA

```{r}
#mRNA

rownames(data.mrna)<- data.mrna$hugoGeneSymbol
```

```{r}
data.mrna<- subset(data.mrna, select = -c(entrezGeneId,type,hugoGeneSymbol))
```

```{r}
data.mrna$variance<- apply(data.mrna, 1, var)
```

```{r}
#Test if the dataset has samples with variance < 0.02

data.mrna_TEST<- subset(data.mrna, variance < 0.02)
```

```{r}
#Clean genes where variance is below 0.02

data.mrna<- subset(data.mrna, variance >= 0.02)
```

Variance filter - miRNA

```{r}
#miRNA

rownames(data.mirna)<- data.mirna$hugoGeneSymbol 
```

```{r}
data.mirna<- subset(data.mirna, select = -c(entrezGeneId,type,hugoGeneSymbol )) 
```

```{r}
data.mirna$variance<- apply(data.mirna, 1, var)
```

```{r}
#Test if the dataset has samples with variance < 0.02

data.mirna_TEST<- subset(data.mirna, variance<0.02)
```

```{r}
#Clean miRNA where variance is below 0.02

data.mirna<- subset(data.mirna, variance>=0.02)
```

Transpose the datasets

```{r}
#miRNA
data.mirna<-t(data.mirna)
```

```{r}
#mRNA
data.mrna<-t(data.mrna)
```

```{r}
#DEFINE A SEED

set.seed(2909)
```

#TRAIN-TEST-SPLIT 70/30- BALANCED

##Clinical: Train -Test Split (70/30)

```{r}
table(data.clinical$PAM50_SUBTYPE)
```

```{r}
data.clinical.Basal<- filter(data.clinical, data.clinical$PAM50_SUBTYPE == 'Basal-like')
data.clinical.Her2<- filter(data.clinical, data.clinical$PAM50_SUBTYPE == 'HER2-enriched')
data.clinical.LumA<- filter(data.clinical, data.clinical$PAM50_SUBTYPE == 'Luminal A')
data.clinical.LumB<- filter(data.clinical, data.clinical$PAM50_SUBTYPE == 'Luminal B')
```

```{r}
#Basal-like

split<- sample.split(data.clinical.Basal, SplitRatio = 0.7)

train.clinical.Basal<- subset(data.clinical.Basal, split==TRUE)
test.clinical.Basal<- subset(data.clinical.Basal, split==FALSE)
```

```{r}
#Her2

split<- sample.split(data.clinical.Her2, SplitRatio = 0.7)

train.clinical.Her2<- subset(data.clinical.Her2, split==TRUE)
test.clinical.Her2<- subset(data.clinical.Her2, split==FALSE)
```

```{r}
#LumA

split<- sample.split(data.clinical.LumA, SplitRatio = 0.7)

train.clinical.LumA<- subset(data.clinical.LumA, split==TRUE)
test.clinical.LumA<- subset(data.clinical.LumA, split==FALSE)
```

```{r}
#LumB

split<- sample.split(data.clinical.LumB, SplitRatio = 0.7)

train.clinical.LumB<- subset(data.clinical.LumB, split==TRUE)
test.clinical.LumB<- subset(data.clinical.LumB, split==FALSE)
```

```{r}
#Make as list

df.list.train<- list(train.clinical.Basal, train.clinical.Her2, train.clinical.LumA, train.clinical.LumB)
df.list.test<- list(test.clinical.Basal, test.clinical.Her2, test.clinical.LumA, test.clinical.LumB)
```

```{r}
#Merge

train.clinical<- Reduce(function(x, y) merge(x, y, all=TRUE), df.list.train)  
test.clinical<- Reduce(function(x, y) merge(x, y, all=TRUE), df.list.test)  
```

```{r}
dim(train.clinical)
```

```{r}
sampleId.train<- train.clinical$sampleId
sampleId.test<- test.clinical$sampleId
```

##mRNA: Train -Test Split (70/30)

```{r}
train.mrna<- data.mrna[rownames(data.mrna) %in% sampleId.train,]
test.mrna<- data.mrna[rownames(data.mrna) %in% sampleId.test,]
```

```{r}
dim(train.mrna)
```

##miRNA: Train -Test Split (70/30)

```{r}
train.mirna<- data.mirna[rownames(data.mirna) %in% sampleId.train,]
test.mirna<- data.mirna[rownames(data.mirna) %in% sampleId.test,]
```

```{r}
dim(train.mirna)
```

## miRNA and mRNA datasets filter

```{r}
#Filter to  have the correspondent samples in both mRNA and miRNA

train.mrna<- train.mrna[rownames(train.mrna) %in% rownames(train.mirna) ,]

test.mrna<- test.mrna[rownames(test.mrna) %in% rownames(test.mirna) ,]
```

```{r}
dim(train.mrna)
```

```{r}
dim(test.mrna)
```

```{r}
#Filter to  have the correspondent samples in mRNA miRNA and Clinical

train.clinical<- train.clinical[train.clinical$sampleId %in% rownames(train.mirna),]

test.clinical<- test.clinical[test.clinical$sampleId %in% rownames(test.mirna),]
```

```{r}
dim(train.clinical)
```

```{r}
dim(test.clinical)
```

##Train-Test Datasets

```{r}
#Train data
train.data<- list(mirna.train = train.mirna, mrna.train = train.mrna, clinical.train = train.clinical)

#Test data
test.data<- list(mirna.test = test.mirna, mrna.test = test.mrna, clinical.test = test.clinical)
```

```{r}
data = list(miRNA= train.data$mirna.train, mRNA = train.data$mrna.train)

lapply(data, dim)
```

```{r}
#Convert to factor as in the dataset from the tutorial

Y<- as.factor(train.data$clinical.train$PAM50_SUBTYPE)
```

```{r}
summary(Y)
```

#Design-matrix with Pairwise PLS

```{r}
pls <- spls(data[["miRNA"]], data[["mRNA"]])
```

```{r}
correlation<- cor(pls$variates$X, pls$variates$Y)
colnames(correlation)<- c("miRNA", "mRNA")
rownames(correlation)<- c("miRNA", "mRNA")
```

```{r}
correlation
```

#Features Tunning
```{r}
test.keepX = list(mRNA = c(50, 60, 100, 120, 130), miRNA = c(50, 60, 100, 120, 130))
```

```{r}
ncomp = 2
```

```{r}
tune.TCGA = tune.block.splsda(X = data,Y = Y, ncomp= ncomp,test.keepX = test.keepX,design = correlation,validation = 'Mfold',folds = 5, nrepeat = 3,dist = "centroids.dist", progressBar = T)
```

```{r}
list.keepX = tune.TCGA$choice.keepX
```

```{r}
list.keepX
```

```{r}
#Tunned features

list.keepX$mRNA<- c(120, 120)
list.keepX$miRNA<- c(130, 50)

list.keepX
```

```{r}
#Optimized design-matrix

design = matrix(c(0.0, 0.7, 0.7), 2, 2)
rownames(design) = c('miRNA', 'mRNA')
colnames(design) = c('miRNA', 'mRNA')
```

```{r}
design
```

#MODEL TRAINING
```{r}
final.diablo.model = block.splsda(X = data, Y = Y, ncomp = ncomp, keepX = list.keepX, design = design)
```

```{r}
final.diablo.model$design
```

```{r}
final.diablo.model
```

#MODEL TESTING
```{r}
data.test = list(mRNA = test.data$mrna.test, miRNA = test.data$mirna.test)
```

```{r}
#Save test data
saveRDS(data.test, file ='data.test-Ari')

#Save test data w/predictions
saveRDS(test.data, file ='data.test.pred-Ari')
```

```{r}
pred = predict(final.diablo.model, newdata = data.test)
```

```{r}
confusion_matrix = get.confusion_matrix(truth = test.data$clinical.test$PAM50_SUBTYPE, predicted = pred$WeightedVote$centroids.dist[,2])
```

```{r}
confusion_matrix
```

#PERFORMANCE EVALUATION

```{r}
evaluation_metrics <- function(confusion_mat){
  
  tp_Bsl<- confusion_mat['Basal-like', 'predicted.as.Basal-like']
  tp_Her2<- confusion_mat['HER2-enriched', 'predicted.as.HER2-enriched']
  tp_LumA<- confusion_mat['Luminal A', 'predicted.as.Luminal A']
  tp_LumB<- confusion_mat['Luminal B', 'predicted.as.Luminal B']
  
  fp_Bsl<- sum(confusion_mat[, 'predicted.as.Basal-like']) - tp_Bsl
  fp_Her2<- sum(confusion_mat[, 'predicted.as.HER2-enriched']) - tp_Her2
  fp_LumA<- sum(confusion_mat[, 'predicted.as.Luminal A']) - tp_LumA
  fp_LumB<- sum(confusion_mat[, 'predicted.as.Luminal B']) - tp_LumB
  
  fn_Bsl<- sum(confusion_mat['Basal-like', ]) - tp_Bsl
  fn_Her2<- sum(confusion_mat['HER2-enriched', ]) -  tp_Her2
  fn_LumA<- sum(confusion_mat['Luminal A', ]) - tp_LumA
  fn_LumB<- sum(confusion_mat['Luminal B', ]) - tp_LumB
  
  total = sum(confusion_mat)
  
  tn_Bsl<- total - (tp_Bsl + fp_Bsl + fn_Bsl)
  tn_Her2<- total - (tp_Her2 + fp_Her2 + fn_Her2) 
  tn_LumA<- total - (tp_LumA + fp_LumA + fn_LumA)
  tn_LumB<- total - (tp_LumB + fp_LumB + fn_LumB)
  
  tp<- sum(tp_Bsl, tp_Her2, tp_LumA, tp_LumB)
  tn<- sum(tn_Bsl, tn_Her2, tn_LumA, tn_LumB)
  fp<- sum(fp_Bsl, fp_Her2, fp_LumA, fp_LumB)
  fn<- sum(fn_Bsl, fn_Her2, fn_LumA, fn_LumB)
  
  
  # Calculate accuracy
  accuracy <- (tp+tn) / (tp + tn + fp + fn)
  
  # Calculate precision
  precision <- tp / (tp+fp)
  
  # Calculate recall
  recall <- tp / (tp+fn)
  
  # Calculate F1
  f1 <- 2 * (precision*recall) / (precision+recall)
  
  metrics <- list( 'accuracy' = accuracy,
                   'precision' = precision,
                   'recall' = recall,
                   'F1' = f1)
  return(metrics)
}
```

```{r}
evaluation_metrics(confusion_matrix)
```

```{r}
get.BER(confusion_matrix)
```

#Save the model

```{r}
saveRDS(final.diablo.model, file = 'Model-Ari2')
```

#PLOTS

```{r}
#ROC curve- miRNA

auc.splsda = auroc(final.diablo.model, roc.block = "miRNA", 
                   roc.comp = 2, print = FALSE)
```

```{r}
#ROC curve- mRNA

auc.splsda = auroc(final.diablo.model, roc.block = "mRNA", 
                   roc.comp = 2, print = FALSE)
```

```{r}
#PlotIndiv

plotIndiv(final.diablo.model, ind.names = FALSE, legend = TRUE, 
          title = 'DIABLO Plots')
```

```{r}
#Loadings- Component 1

png('Graph_Load_Comp1.png', width = 10, height =20, units = 'in', res= 300)
plotLoadings(final.diablo.model, comp = 1, contrib = 'max', method = 'median')
dev.off()
```

```{r}
#Loadings- Component 2

plotLoadings(final.diablo.model, comp = 2, contrib = 'max', method = 'median')
```


```{r}
#Circus Plot
circosPlot(final.diablo.model, cutoff = 0.6, line = TRUE,
           color.blocks= c('darkorchid', 'brown1'),
           color.cor = c("blue","red"), size.labels = 1.5)
```

```{r}
#Heatmap
cimDiablo(final.diablo.model)
```